Estimating duration in the presence of misclassification

### 2024-01-28 incidence-prevalence.tex
editing

To relax the assumption of a homogeneous population, the population can be stratified.
If the strata have differing transmission patterns, this can change the dynamics of the epidemic.
The different patterns could be due to behavioural factors (\eg high-numbers of sexual partners in a sexually-transmitted disease~\autocite[69]{keelingModeling} or younger age groups making many more contacts for respiratory diseases~\autocite[176]{andersonInfectious}), or biological factors such as genetics~\autocite[208]{andersonInfectious}..
I denote the set of strata, as $\set{A}$, with the set's size as $A$.
I will assume that the disease progression (\ie length of the exposed and infectious periods) is the same for all strata, although this assumption can be relaxed.
Stratification requires two major  model changes: extension of the state vector and a description of the between-strata interactions.

>>>>

To relax the assumption of a homogeneous population, the population can be stratified.
If the strata have differing transmission patterns, this can change the dynamics of the epidemic.
The different patterns could be due to behavioural factors (\eg high-numbers of sexual partners in a sexually-transmitted disease~\autocite[69]{keelingModeling} or younger age groups making many more contacts for respiratory diseases~\autocite[176]{andersonInfectious}), or biological factors such as genetics~\autocite[208]{andersonInfectious}.
I denote the set of strata, as $\set{A}$, with the set's size as $A$.
I will assume that the disease progression (\ie length of the exposed and infectious periods) is the same for all strata, although this assumption can be relaxed.
Stratification requires two major  model changes: extension of the state vector and a description of the between-strata interactions.

### 2024-01-28 transmission.tex
editing

\todo[inline]{Think about the framing of this paragraph, maybe talk about being linear in its parameters and pre-existing software packages.}
The advantage of fitting a multilevel regression model on prevalence, rather than incidence, is that prevalence with appropriate prior (or penalisation) can be expressed as a generalised linear model (GLM)\todo{maybe cite backcalculation as a GLM}.
Either incidence proportion or prevalence, $pi_i$ in a stratum $i$ can be expressed as $\logit(\pi_i) = g(i, t)$, where is a function of the stratum and time, penalised towards being linear in time.
If $g$ is chosen appropriately, this is a GLM with a logit link function.

@@

Having fit the prevalence model, I draw 10,000\todo{check number} prevalence trajectories in each stratum from the approximate posterior distribution of the model.
For each stratum, I use the relationship in \cref{E-inc-prev:eq:EPt} to calculate the incidence proportion at each time point.
I consider the relationship between incidence and prevalence within a stratum to be deterministic, as \cref{E-inc-prev:sec:prevalence-process} justifies, that is $P_t = \E(P_t \mid Z)$.
\Cref{E-inc-prev:eq:EPt}, considered across different values of $t$, can be written as $\vec{P} = \matr{S} \vec{Z}$, where: $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum; $\matr{S}$ is a matrix formed from the survival function, as defined below; and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
The matrix $\matr{S}$ is defined to maintain the relationship between incidence and prevalence.
This gives a lower-triangular of the following form:
\begin{align}
    (\matr{S})_{i,j} = \begin{cases}
        S(i - j + 1) & i - \dmax - 1 \leq j \leq i \\
        0 & \text{otherwise.}
    \end{cases}
\end{align}
Since the matrix is lower-triangular, the solution for $\vec{P}$ can be found efficiently using forward substitution\todo{cite forward substitution}.
Forward substitution is numerically stable unless the diagonal elements are close to 0.
Here, the diagonal is $S(1) = 1$ and hence the matrix is well-conditioned.
\todo{I explained this more clearly in my first-year report, look that up!}

@@

The final step is poststratifying the incidence proportion to the CIS target population and subpopulations of interest.
For this, I first draw 500 samples from the approximate posterior distribution estimated.
For each posterior draw and day, the incidence proportion in the (sub)population is the weighted average of the incidence proportion in each stratum, where the weights are the proportion of the population in each stratum.
The population numbers for each stratum are provided by the Office for National Statistics.

@@

During November, incidence starts rising, although starting at different times in different regions.
This could be due to the emergence of the Alpha variant, which was more transmissible than previous variants and emerged at different times across the country\todo{cite Alpha variant}.
A second peak occurs in late December or early January in all regions; this second peak is larger than the first in all regions except the Northern three (North East, North West, and Yorkshire and the Humber).
\begin{figure}
    \includegraphics{transmission/backcalc-regions}
    \caption[Incidence estimated using backcalculation by region]{%
        Incidence proportion for all of England (A) and its component regions.
        Regions are split into broad groupings of three regions each: Northern (B), Midlands (C), and Southern (D).
        Yorkshire and the Humber is labelled as Yorkshire for brevity.
    }
    \label{transmission:fig:backcalc-regions}
\end{figure}

@@

\subsection{Methods}
\todo[inline]{add a bit more intro in the context of chapter 3}
The model will be stratified only by region and age group, so the data are aggregated to these levels.
Only individuals aged two or older are eligible to participate in the CIS; I assume that test results in all participants aged under 11 are still representative of the whole $[0, 11)$ age group.

As well as the motivation given in \cref{E-backcalc}, independent regions allows heterogeneity in the parameters between regions, which could arise from a range of socioeconomic factors.
Furthermore, the assumption of homogeneous mixing within age groups is more plausible for regions than the whole of England.
By modelling each region independently, I assume that the epidemics in each region do not interact.
Clearly, this is not realistic as travel between regions of England is common and unrestricted.
However, for a large epidemic, the non-interaction is a good approximation~\autocite{birrellRealtimea}.

@@

A parsimonious model is needed to parameterise the WAIFW matrix (see \cref{E-SEIR:sec:structured-populations}).
I follow the WAIFW matrix parameterisation of \textcite{birrellRealtime}, decomposition $\beta_{aa'}(t) = \beta(t) \kappa_{aa'}(t) q_{aa'} N_{a'}$.\todo{remind readers what these quantities are here}
This separates contact patterns, $\kappa_{aa'}$, which can vary over time; transmission probability upon contact between age groups $a$ and $a'$, $q_{aa'}$, assumed constant; a region-wide effect for time, $\beta(t)$; and the population of the stratum, $N_{a'}$.

@@

\Cref{transmission:fig:compare-regions} compares the regional incidence estimates between the two models in this chapter.
Generally the estimates agree, with the SEIR model having much narrower credible intervals.
The largest discrepancy is that the backcalculation model estimates a second peak in mid January in several regions (most clearly in London, North West, and the East of England).
However, except in London, the credible intervals still overlap.
This is a specific case of a larger pattern: the SEIR model produces smoother estimates of incidence than the backcalculation model.
\begin{figure}
    \centering \includegraphics{transmission/compare-regions}
    \caption[Comparing each models estimate of regional incidence.]{%
        Incidence proportion in England and each region estimated by the two models discussed in this chapter.
        Posterior median and 95\% CrI shown.
        For the SEIR model, newly PCR positive (entrances into the P1 compartment), rather than incidence (entrances into the E compartment) are shown for comparability with the backcalculation results.
    }
    \label{transmission:fig:compare-regions}
\end{figure}

@@

\todo[inline]{write conclusion and add to discussion once comparison with backcalculation is made}

>>>>

\todo[inline]{Think about the framing of this paragraph, maybe talk about being linear in its parameters and pre-existing software packages.}
The advantage of fitting a multilevel regression model on prevalence, rather than incidence, is that prevalence with appropriate prior (or penalisation) can be expressed as a generalised linear model (GLM)\todo{maybe cite backcalculation as a GLM}.
Either incidence proportion or prevalence, $\pi_i$ in a stratum $i$ can be expressed as $\logit(\pi_i) = g(i, t)$, where is a function of the stratum and time, penalised towards being linear in time.
If $g$ is chosen appropriately, this is a GLM with a logit link function.

@@

Having fit the prevalence model, I draw a sample of 500 prevalence time series in each stratum from the approximate posterior distribution.
For each stratum, I use the relationship in \cref{E-inc-prev:eq:EPt} to calculate the incidence proportion at each time point.
I consider the relationship between incidence and prevalence within a stratum to be deterministic, as \cref{E-inc-prev:sec:prevalence-process} justifies, that is $P_t = \E(P_t \mid Z)$.
\Cref{E-inc-prev:eq:EPt}, considered across different values of $t$, can be written as $\vec{P} = \matr{S} \vec{Z}$, where: $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum; $\matr{S}$ is a matrix formed from the survival function, as defined below; and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
The matrix $\matr{S}$ is defined to maintain the relationship between incidence and prevalence.
This gives a lower-triangular of the following form:
\begin{align}
    (\matr{S})_{i,j} = \begin{cases}
        S(i - j + 1) & i - \dmax - 1 \leq j \leq i \\
        0 & \text{otherwise.}
    \end{cases}
\end{align}
$\matr{S}$ is lower-triangular, hence to solve for $\vec{P}$ I use forward-substitution~\autocite{cormenMatrix}.
Forward substitution is numerically stable unless the diagonal elements, here $S(1)$, are small relative to the off-diagonal elements, here $S(i)$ for $i > 0$~\autocite[section 8.2]{highamAccuracy}.
Here, the diagonal is $S(1) = 1$ and hence the matrix the solution is numerically stable.

@@

The final step is poststratifying the incidence proportion to the CIS target population and subpopulations of interest.
For this, I start with the 500 samples of the approximate posterior distribution of the incidence.
For each posterior draw and day, the incidence proportion in the (sub)population is the weighted average of the incidence proportion in each stratum, where the weights are the proportion of the population in each stratum.
The population numbers for each stratum are provided by the Office for National Statistics.

@@

During November, incidence starts rising, although starting at different times in different regions.
This could be due to the emergence of the Alpha variant, which was more transmissible than previous variants and emerged at different times across the country~\autocites{walkerTracking}{daviesEstimated}{lythgoeLineage}.
A second peak occurs in late December or early January in all regions; this second peak is larger than the first in all regions except the Northern three (North East, North West, and Yorkshire and the Humber).
\begin{figure}
    \includegraphics{transmission/backcalc-regions}
    \caption[Incidence estimated using backcalculation by region]{%
        Incidence proportion for all of England (A) and its component regions.
        Regions are split into broad groupings of three regions each: Northern (B), Midlands (C), and Southern (D).
        Yorkshire and the Humber is labelled as Yorkshire for brevity.
    }
    \label{transmission:fig:backcalc-regions}
\end{figure}

@@

\subsection{Methods}

% The model will be stratified only by region and age group, so the data are aggregated to these levels.
% Only individuals aged two or older are eligible to participate in the CIS; I assume that test results in all participants aged under 11 are still representative of the whole $[0, 11)$ age group.

I model each region fully independently, \ie with no shared parameters or transmission between regions.
As well as the motivation given in \cref{E-backcalc}, independent regions allows heterogeneity in the parameters between regions, which could arise from a range of socioeconomic factors.
Furthermore, the assumption of homogeneous mixing within age groups is more plausible for regions than the whole of England.
By modelling each region independently, I assume that the epidemics in each region do not interact.
Clearly, this is not realistic as travel between regions of England is common and unrestricted.
However, for a large epidemic, the non-interaction is a good approximation~\autocite{birrellRealtimea}.

@@

A parsimonious model is needed to parameterise the WAIFW matrix (see \cref{E-SEIR:sec:structured-populations}).
I follow the WAIFW matrix parameterisation of \textcite{birrellRealtime}, decomposing the rate of transmission from individuals in age group $a'$ to susceptible individuals in age group $a$ as $\beta_{aa'}(t) = \beta(t) \kappa_{aa'}(t) q_{aa'} N_{a'}$.
This separates contact patterns, $\kappa_{aa'}$, which can vary over time; transmission probability upon contact between age groups $a$ and $a'$, $q_{aa'}$, assumed constant; a region-wide effect for time, $\beta(t)$; and the population of the stratum, $N_{a'}$.

@@

\Cref{transmission:fig:compare-regions} compares the regional incidence estimates between the two models in this chapter.
Generally the estimates agree, with the SEIR model having much narrower credible intervals.
The largest discrepancy is that the backcalculation model estimates a second peak in mid January in several regions (most clearly in London, North West, and the East of England).
However, except in London, the credible intervals still overlap.
This is a specific case of a larger pattern: the SEIR model produces smoother estimates of incidence than the backcalculation model.;
\begin{figure}
    \centering \includegraphics{transmission/compare-regions}
    \caption[Comparing each models estimate of regional incidence.]{%
        Incidence proportion in England and each region estimated by the two models discussed in this chapter.
        Posterior median and 95\% CrI shown.
        For the SEIR model, newly PCR positive (entrances into the P1 compartment), rather than incidence (entrances into the E compartment) are shown for comparability with the backcalculation results.
    }
    \label{transmission:fig:compare-regions}
\end{figure}

@@

The major contribution in this chapter is showing that transmission and incidence can be estimated using only a
